---
title: "Individual Assignment 5"
author: "Natalie Chen"
date: 05-04-2026
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute: 
  warning: false
---

# Set up

## Packages
```{r}
#| label: packages

#general use 
library(here)

library(tidyverse)

library(janitor)

library(snakecase)

# packages for visualization customization

library(scales)
library(paletteer)
library(ggthemes)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)

```

## Data
```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```



# 1. Visualizing differences across sites in major taxon abundance

## a. Plan your figure

- x-axis: site
- y-axis: abundance of ostracods
- geometry to represent average abundance/L: `geom_jitter()`
- geometry to represent medians: `stat_summary()`


## b. Wrangle your data

```{r}
#| label: wrangling ostracod

# creating new object from aquatic insects 
ostracod <- aq_ins_clean |> 
  
  # filter to only include ostracods
  filter(taxon == "ostracod")

# display 10 random rows

slice_sample(ostracod, n = 10)

```




## c. Code your figure


## d. Create a multipanel figure


## e. Interpret your figure


# 2. Visualizing total abundan ce as a function of salinity

## a. Plan your figure

## b. Wrangle your data

## c. Code your figure

## d. Interpret your figure
