---
title: "Individual Assignment 5"
author: "Natalie Chen"
date: 05-04-2026
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
execute: 
  warning: false
---

# Set up

## Packages
```{r}
#| label: packages

#general use 
library(here)

library(tidyverse)

library(janitor)

library(snakecase)

# packages for visualization customization

library(scales)
library(paletteer)
library(ggthemes)

# reading in .xlsx files
library(readxl)

# visualizing missing data
library(naniar)

# arranging plots
library(patchwork)

```

## Data
```{r}
#| label: data

# taxonomic information
taxon_list <- read_csv(here("data", "taxon_list.csv")) 

# aquatic invertebrates
aq_ins <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                    sheet = "Aquatic Insects") 

# site information
sites <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                   sheet = "sites")

# water quality information
water_quality <- read_xlsx(here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
                           sheet = "Water Quality",
                           na = c("", "n/a", "N/A", "over", "173+")) 
```

**copied from class**

```{r}
#| label: ncos sites

# creating new object from sites
ncos_sites <- sites |> 
  
  # filter to only include NCOS quarterly sampling sites
  filter(sector == "NCOS" & sampling_frequency == "Quarterly")

```

```{r}
#| label: cleaning-water-quality

water_quality_clean <- water_quality |> 
  
  # cleaning column names
  clean_names() |> 
  
  # filtering join: only include sites in ncos_sites
  # keep all the rows on the left that match up with rows on the right
  semi_join(ncos_sites, by = "site") |> 
  
  # combine date and site together and remove those indiv. columns
  unite("date_site", date, site, remove = TRUE) |> 
  
  # group by date_site
  group_by(date_site) |> 
  
  # create new columns that show median of each variable at each date_site
  # because there are multiple samples at the same site at different depths
  summarize(med_ph = median(p_h, na.rm = TRUE),
            med_sal = median(salinity_ppt, na.rm = TRUE),
            med_do = median(dissolved_oxygen_mg_l, na.rm = TRUE)) |> 
  
  # ungroup
  ungroup() |> 
  
  # filter out the salinity outlier
  filter(date_site != "2022-08-12_NVBR")


```

```{r}
#| label: cleaning-taxon

# creating new object from taxon_list
taxon_list_clean <- taxon_list |> 
  
  # converted taxon name values to snake case (lowercase)
  mutate(verbatim_name = to_any_case(verbatim_name,
                                     case = "snake"))

```

```{r}
#| label: cleaning-aq-ins

# creating new clean aquatic inverts object from aquatic inverts
aq_ins_clean <- aq_ins |> 
  
  # cleaning column names
  clean_names() |> 
  
  # filtering join: only include quarterly NCOS surveys from ncos_sites
  semi_join(ncos_sites, by = "site") |> 
  
  # renaming column to just date
  rename(date = date_on_vial) |> 
  
  # select columns of interest
  select(date, sample_type, site,
         
         # 9 most abundant taxon
         ostracod,
         copepod,
         hemiptera_corixidae_boatman,
         cladocera,
         nematode,
         diptera_ceratopogonidae, 
         annelida_oligochaete,
         diptera_chironomid,
         annelida_polychaete) |>
  
  # filter to only include filtered beaker samples 
  filter(sample_type %in% c("FB250", "FB 250"))  |> 
  
  
  pivot_longer(
    # cols = columns to make longer
    cols = ostracod:annelida_polychaete,
    # new column for the old column names
    names_to = "taxon",
    # new column for the values in each of the old columns
    values_to = "abundance" 
  ) |> 
  
  # group by the date, site, and taxon
  group_by(date, site, taxon) |> 
  
  # calculate average abundance per liter (sum abundance divided by 7.5 liters)
  summarize(avg_per_lit = sum(abundance, na.rm = TRUE)/7.5) |> 
  
  # ungroup
  ungroup() |> 
  
   # create a new column called `date_site` to join with water quality
   # don't remove the original columns
  unite("date_site", date, site, remove = FALSE) |> 
  
  # join with water quality clean
  # wqc has `date_site` column as well
  left_join(water_quality_clean, by = "date_site") |> 
  
  # join with taxonomic information from taxon list
  # verbatim_name column in taxon_list_clean matches with taxon column in aq_ins_clean
  left_join(taxon_list_clean, by = c("taxon" = "verbatim_name")) |> 

  # fill in full site names
  mutate(site_full = case_when(
    site == "NVBR" ~ "North of Venoco Bridge",
    site == "NEC" ~ "East Channel",
    site == "NMC" ~ "Main Channel",
    site == "NPB" ~ "South of Phelps Bridge",
    site == "NPB1" ~ "North of Phelps Bridge",
    site == "NPB2" ~ "Phelps Road",
    site == "NWP" ~  "West Pond",
    site == "NDC" ~ "Devereux Creek"
  )) |> 
  
  # setting factor levels for full site name by median salinity
  mutate(site_full = fct_reorder(site_full, med_sal, .fun = "median", .na_rm = TRUE),
         site_full = fct_rev(site_full))


```


```{r}
#| label: visualizing salinity plot
#| fig-height: 4
#| fig-width: 8

salinity_plot <- ggplot(data = aq_ins_clean,
                        # x-axis: full site name
                        # y-axis: median salinity
                        mapping = aes(x = site_full,
                                      y = med_sal)) +
  # points representing median salinity at each survey
  geom_jitter(height = 0,
              shape = 21) +
  
  stat_summary(geom = "point",
               fun = median,
               color = "red",
               size = 2,) +
  
  # wrap x-axis labels
  scale_x_discrete(labels = label_wrap(width = 10)) +
  
  # labeling titles and axes
  labs(x = "Site",
       y = "Median salinity (ppt)",
       title = "Site Salinity") +
  
  # change theme
  theme_bw()


# displaying the object
salinity_plot

```



```{r}
#| label: copepod cleaning

copepods <- aq_ins_clean |> 
  
  filter(taxon == "copepod") |> 
  
  mutate(log_avg_per_lit = log(avg_per_lit + 1))


```


```{r}
#| label: copepod plot
#| fig-height: 4
#| fig-width: 8

# creating object to store plot
copepod_plot <- ggplot(data = copepods,
                       # x-axis: full site name
                       # y-axis: log transformed avg count/L
                       mapping = aes(x = site_full,
                                   y = log_avg_per_lit)) +
  # jitter to represent salinity
  geom_jitter(height = 0,
              width = 0.1, 
              shape = 21) +
# points to represent median copepod abundance
  stat_summary(geom = "point",
               fun = median,
               color = "blue",
               size = 2) +
  # wrapping x axis labels
scale_x_discrete(labels = label_wrap(width = 10)) +
  
  # relabelling axes
  labs(x = "Site",
       y = "log + 1 transformed averag abundance/L",
       title = "Copepod Abundance") +
  
  # different theme
  theme_bw()
  
# display object
copepod_plot

```

# 1. Visualizing differences across sites in major taxon abundance

## a. Plan your figure

- x-axis: site_full
- y-axis: average abundance of ostracod per liter
- geometry to represent average abundance/L: `geom_jitter()`
- geometry to represent medians: `stat_summary()`


## b. Wrangle your data

```{r}
#| label: wrangling ostracod

# creating new object from aquatic insects 
ostracod <- aq_ins_clean |> 
  
  # filter to only include ostracod
  filter(taxon == "ostracod") |> 
  
  # calculate the average of ostracod abundance per lit
  mutate(ostracod_log_avg_per_lit = log(avg_per_lit + 1))

# display 10 random rows

slice_sample(ostracod, n = 10)

```


## c. Code your figure

```{r}
#| label: visualizing ostracod abundance
#| fig-height: 4
#| fig-width: 8

#  base layer: create an object that stores the plot
ostracod_plot <- ggplot(data = ostracod,
                        # x = site
                        # y = log + 1 transformed average abundance/L
                        mapping = aes(x = site_full,
                                      y = ostracod_log_avg_per_lit)) +
  
  # first layer: geometry jitter points to represent each sample 
  geom_jitter(height = 0,
              width = 0.1,
              shape = 21) +
  # points to represent median abundance
  stat_summary(geom = "point",
               fun = median,
               color = "green",
               size = 2) +
  # wrapping x axis labels
  scale_x_discrete(labels = label_wrap(width = 10)) +
  
  # labeling axes and title
  labs(title = "Ostracod Abundance",
       x = "Site",
       y = "log + 1 transformed average abundance/L") +
  
  # changed theme
  theme_bw()

# display plot
ostracod_plot

```

## d. Create a multipanel figure

```{r}
#| label: multipanel salinity/copepod/ostracod figure
#| fig-height: 10
#| fig-width: 10

salinity_plot / (copepod_plot | ostracod_plot)


```


## e. Interpret your figure

**What are the differences between sites in copepod and ostracod abundance?**

- Average copepod abundance/L was highest at East Channel and Devereux Creek at respective transformed values of around 4 and 3, and moderate at a value of around 2 at North of Venoco Bridge, South of Phelps Bridge, and Phelps Road. The highest average ostracod abundance/L was moderate around a transformed value of 2 at South of Phelps Bridge and also Devereux Creek, and low at pretty much every other site.

**How does ostracod abundance relate (or not relate) to site-level differences in salinity?**

- Ostracod abundance does not correlate to salinity site-level differences in salinity because the plots do not show the same type of trend for the salinity plot and for the ostracod abundance plot. 

# 2. Visualizing total abundance as a function of salinity

## a. Plan your figure

- x-axis: salinity (ppt)
- y-axis: log-transformed average abundance/L
- color: taxon
- geometry: geom_point()
- facets: by each taxon in aq_ins_clean with free scales 


## b. Wrangle your data


```{r}
#| label: wrangling aq insects


# change existing aq_in_clean object
aq_ins_clean <- aq_ins_clean |> 

  # group by the sample date and by taxon
  group_by(date, taxon) |> 
  
  # add in a new column that is the sum of avg_per_lit per taxon per date
  mutate(total_avg_per_lit = sum(avg_per_lit, na.rm = TRUE)) |> 
  
  # ungroup
  ungroup() |> 
  
  # keep only date, taxon, med_sal, and total avg/L
  select(date, taxon, med_sal, total_avg_per_lit) |> 

  # reorder by max abundance
  arrange(desc(total_avg_per_lit))
  

# display 10 random rows
slice_sample(aq_ins_clean, n = 10)

```

## c. Code your figure



## d. Interpret your figure
